---
title: "Case Study 1 - Group 1"
author:
- Annika Janson h11829506
- Jan Beck h11814291
- Franz Uchatzi h1451890
date: "29.11.2020"
output:
  pdf_document: default
  html_document:
    df.print: paged
  word_document: default
header-includes:
- \usepackage{dcolumn}
- \renewcommand{\and}{\\}
---


```{r setup, include=FALSE}
library(car)
library(stargazer)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
gdp <- read.csv("USGDP.csv")
N <- nrow(gdp)
```

# 1.2 Tasks

```{r, echo=FALSE}

ts_gdp <- ts(gdp$NA000334Q/10^6, frequency = 4, start = c(2010, 1)) # convert to trillions
plot(ts_gdp, main = "US GDP (Q1 2010 - Q4 2020)", xaxs="i", yaxs="i", las=1, ylab = "USD, in trillions", xlab = "Year", ylim=c(3.5, 6), xlim=c(2010,2021), frame.plot=FALSE, col="red", axes=FALSE)
axis(side=1, at=c(2010:2021), las=2) # custom x-axis 
axis(side=2, at=seq(3.5, 6, by=.5), las=2) # custom y-axis
grid(NA, 5, col = "lightgray") # grid only in y-direction


```
In the chart above we plotted US gross domestic product (GDP) from the first quarter of 2010 till the fourth quarter of 2020 in trillions of Dollars. We can see a general positive trend up until 2020 - the year of the outbreak of the coronavirus pandemic. The minimum in the observed time frame was USD ~3.5 trillion in 2010 and the maximum was USD ~5.5 trillion in 2019. There appears to be a seasonal trend in every first quarter of the year, which is generally lower than the previous quarter.

```{r, echo=FALSE}

ts_gdp_log <- ts(log(gdp$NA000334Q), frequency = 4, start = c(2010, 1)) # convert to trillions
plot(ts_gdp_log, main = "US GDP (Q1 2010 - Q4 2020)", xaxs="i", yaxs="i", las=1, ylab = "USD, logged", xlab = "Year", ylim=c(15,15.6), xlim=c(2010,2021), frame.plot=FALSE, col="red", axes=FALSE)
axis(side=1, at=c(2010:2021), las=2) # custom x-axis 
axis(side=2, at=seq(15, 16.6, by=.1), las=2) # custom y-axis
grid(NA, 6, col = "lightgray") # grid only in y-direction

```
(Sehe nicht wirklich einen Unterschied - oder hab einen Fehler gemacht?)

\newpage

### 1.2.1

### 1.2.2

### 1.2.3

### 1.2.4

### 1.2.5

### 1.2.6

### 1.2.7

### 1.2.8

## 2 Tasks

### 1

### 2